Determinism, Complexity, and Predictability in Computer Performance

### 2014-10-15 FinalSubmissiontoPRE/pecomplexity.tex
fixing some wording for greyscale

% Commented this out because the addition of the paragraph from the
% introduction seemed like it could replace it
% 
% \col is a simple C program that repeatedly initializes the upper
% triangle of a 2048 $\times$ 2048 matrix in column-major order.
% % 
% % by looping over the following three lines of code:
% % \begin{lstlisting}
% %   for(i=0;i<2048;i++)
% %   	for(j=i;j<2048;j++)
% %   		data[j][i] = 0;
% % \end{lstlisting}
% % full listing
% % #define SIZE (2048)
% % 
% % int main(int argc, char ** argv)
% % {
% %   int i, j, n;
% %   static int data[SIZE][SIZE];
% %   
% %   for(n=0;n<numInits;n++){  
% %   for(i=0;i<SIZE;i++)
% %   	for(j=i;j<SIZE;j++)
% %   		data[j][i] = 0;
% %   }
% %   return 0;
% % }
% %
% As mentioned in Section~\ref{sec:intro}, this simple program exhibits
% surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
% of the processor performance during the execution of this program is
% shown in Figure~\ref{fig:col-ts}.
% \begin{figure}[htbp]
%   \centering
%     \includegraphics[width=\columnwidth]{figs/colFullTS}
%     \caption{The instructions executed per CPU clock cycle (IPC) as
%       the \col program runs.  Each point is the average IPC in a
%       100,000 instruction period. Figure~\ref{fig:col-ipc} is a
%       closeup of the first part of this signal.  }
%     \label{fig:col-ts}
%   
%   \end{figure}

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our
experiments, we chose $M=750$ and $N=1000$ and generated the matrix
entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as is clearly visible in Figure
\ref{fig:svd-ts-colored}.  
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{SVD1RegimesColored}
    \caption{(Color online) An IPC trace during the execution of \gcc.  The colors (also separated by vertical dashed lines)
      identify the different \emph{segments} of the signal that are
      discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
As the code moves though its different phases---diagonalizing the
matrix, computing its transpose, multiplying, etc.---the processor
utilization patterns change quite radically.  For the first
$\sim$21,000 measurements (21,000 $\times$ 100,000 instructions),
roughly 1.8 instructions are executed per cycle, on the average, by
the eight processing units on this chip.  After that, the IPC moves
through a number of different oscillatory regimes, which we have
color-coded in the figure in order to make textual cross-references
easy to track.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  This
is a different nonstationarity than in the case of \col---large jump
discontinuities rather than small shifts in the baseline---and not one
that we were able to handle by simply forcing the ARIMA model to
include a particular term.  The \naive ~method has a very difficult
time with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
and in turn raising both the mean and variance of this set of trials.

>>>>

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our
experiments, we chose $M=750$ and $N=1000$ and generated the matrix
entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as is clearly visible in Figure
\ref{fig:svd-ts-colored}.  
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{SVD1RegimesColored}
    \caption{(Color online) An IPC trace during the execution of \gcc.  The colors (also separated by vertical dashed lines)
      identify the different \emph{segments} of the signal that are
      discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
As the code moves though its different phases---diagonalizing the
matrix, computing its transpose, multiplying, etc.---the processor
utilization patterns change quite radically.  For the first
$\sim$21,000 (in units of $10^5$ instructions) measurements,
roughly 1.8 instructions are executed per cycle, on the average, by
the eight processing units on this chip.  After that, the IPC moves
through a number of different oscillatory regimes, which we have
color-coded in the figure (as well as separated by vertical dashed lines) in order to make textual cross-references
easy to track.

@@

\svdone---the dark blue (first) segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  This
is a different nonstationarity than in the case of \col---large jump
discontinuities rather than small shifts in the baseline---and not one
that we were able to handle by simply forcing the ARIMA model to
include a particular term.  The \naive ~method has a very difficult
time with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
and in turn raising both the mean and variance of this set of trials.
